
---
title: "Using tidyr in R"
author: "Ana Jane"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction
The `tidyr` package is a part of the `tidyverse` and is used to make data tidy, which means data should be structured in a way that each variable is a column, each observation is a row, and each value is a cell. This R Markdown document provides an overview of some key functions in the `tidyr` package.

# Load Libraries
```{r setup, include=TRUE}
library(tidyr)
library(dplyr)
```

# Sample Dataset
We will use a small dataset for demonstration purposes:
```{r}
data <- data.frame(
  ID = 1:6,
  Name = c("Alice", "Bob", "Charlie", "Diana", "Eve", "Frank"),
  Math = c(78, 85, 90, 88, 76, 92),
  Science = c(82, 89, 88, 91, 84, 95),
  History = c(70, 75, 80, 85, 72, 90)
)
data
```

# Pivoting Data
## Pivoting from Wide to Long Format
The `pivot_longer()` function transforms the dataset from wide to long format:
```{r}
long_data <- data %>% pivot_longer(
  cols = c(Math, Science, History),
  names_to = "Subject",
  values_to = "Score"
)
long_data
```

## Pivoting from Long to Wide Format
The `pivot_wider()` function reverses the process, transforming the dataset back to wide format:
```{r}
wide_data <- long_data %>% pivot_wider(
  names_from = Subject,
  values_from = Score
)
wide_data
```

# Handling Missing Data
## Creating a Dataset with Missing Values
```{r}
data_with_na <- data.frame(
  ID = 1:6,
  Name = c("Alice", "Bob", "Charlie", "Diana", "Eve", "Frank"),
  Math = c(78, 85, NA, 88, 76, NA),
  Science = c(82, NA, 88, 91, NA, 95)
)
data_with_na
```

## Filling Missing Values
The `fill()` function is used to fill missing values in a dataset:
```{r}
filled_data <- data_with_na %>% fill(Math, .direction = "down")
filled_data
```

## Replacing Missing Values
The `replace_na()` function allows us to replace missing values with specific values:
```{r}
replaced_data <- data_with_na %>% replace_na(list(Math = 0, Science = 0))
replaced_data
```

# Separating and Uniting Columns
## Separating Columns
The `separate()` function splits a column into multiple columns based on a delimiter:
```{r}
separate_data <- data.frame(
  ID = 1:4,
  FullName = c("Alice Johnson", "Bob Smith", "Charlie Brown", "Diana Prince")
) %>% separate(FullName, into = c("FirstName", "LastName"), sep = " ")
separate_data
```

## Uniting Columns
The `unite()` function combines multiple columns into one:
```{r}
united_data <- separate_data %>% unite("FullName", FirstName, LastName, sep = " ")
united_data
```

# Conclusion
The `tidyr` package is essential for data wrangling tasks in R. It provides a variety of tools for reshaping and cleaning data, making it easier to analyze and visualize.
